---
title: "Assignment 04: Analysis of crime in Washington, DC"
author: "Jacob Ausubel"
date: "10/08/23"
format: html
execute: 
  warning: false
urlcolor: blue
---

# Link to dataset

https://opendata.dc.gov/datasets/DCGIS::crime-incidents-in-2021/about

# Topic and motivation

In this analysis, I demonstrate that violent crime is much more common in some parts of Washington, DC than in others. The motivation for the assignment is an uptick in crime in the city in recent years that has received substantial media attention. Lawmakers and policymakers need to keep variation in homicides, sex abuses, assaults with dangerous weapons, and robberies between wards and neighborhood clusters in mind when designing programs aimed at curbing violent crime.

# Initial data setup, cleaning, and manipulation

```{r}
#| warning: false
#| message: false
#Reading in packages
library(tidyverse)
library(readxl)
library(ggthemes)
```

```{r}
#| warning: false
#| message: false
#Removing datasets from working directory
rm(list = ls())
```

```{r}
#| warning: false
#| message: false

#Reading in crime dataset
crime21 <- read.csv("Crime_Incidents_in_2021.csv")

#Creating violent crime variable
#1=Yes, 0=No
crime21$violent_crime <-
  ifelse(crime21$OFFENSE == "HOMICIDE", 1,
    ifelse(crime21$OFFENSE == "SEX ABUSE", 1,
      ifelse(crime21$OFFENSE == "ASSAULT W/DANGEROUS WEAPON", 1, 
        ifelse(crime21$OFFENSE == "ROBBERY", 1, 0))))
```

```{r}
#| warning: false
#| message: false

#Creating a data frame with 8 rows (one for each ward)
#Column for number of violent crimes
violent_crime_by_ward <- crime21 %>%
  filter(violent_crime == 1) %>%
  group_by(WARD) %>%
  summarise(count = n()) %>%
  filter(!is.na(WARD))
```

```{r}
#| warning: false
#| message: false
#Creating a data frame with 32 rows
#Want the counts for violent crimes in each ward broken
  #down by offense type
#8 wards x 4 offense types = 32
violent_crime_by_ward2 <- crime21 %>%
  filter(violent_crime == 1) %>%
  group_by(WARD, OFFENSE) %>%
  summarise(count = n()) %>%
  filter(!is.na(WARD))

#Recoding offense variable
violent_crime_by_ward2$OFFENSE[
  violent_crime_by_ward2$OFFENSE == "ASSAULT W/DANGEROUS WEAPON"
  ] <- "Assault w/dangerous weapon"
violent_crime_by_ward2$OFFENSE[
  violent_crime_by_ward2$OFFENSE == "HOMICIDE"
] <- "Homicide"
violent_crime_by_ward2$OFFENSE[
  violent_crime_by_ward2$OFFENSE == "ROBBERY"
] <- "Robbery"
violent_crime_by_ward2$OFFENSE[
  violent_crime_by_ward2$OFFENSE == "SEX ABUSE"
] <- "Sex Abuse"
```

```{r}
#| warning: false
#| message: false
#Creating a data frame with 32 rows
#Want the counts for violent crimes in each 
  #neighborhood cluster broken
  #down by offense type
#Note that there are 44 neighborhood clusters.
  #Why 154 rows, not 176?
  #After all, 44 x 4 = 176
  #Reason: in some neighborhood clusters, there were
    #0 cases of a specific offense.
violent_crime_by_neigh_cluster <- crime21 %>%
  filter(violent_crime == 1) %>%
  group_by(NEIGHBORHOOD_CLUSTER, OFFENSE) %>%
  summarise(count = n()) %>%
  filter(!is.na(NEIGHBORHOOD_CLUSTER))
```

```{r}
#| warning: false
#| message: false
#Creating a new data frame with 44 rows
#Reason: I want a row for each neighborhood
  #cluster and different columns for counts
  #for each offense type
violent_crime_by_neigh_cluster2 <-
  pivot_wider(violent_crime_by_neigh_cluster,
  names_from = OFFENSE,
  values_from = count)

#Replacing NAs with 0s
#Explanation: in some neighborhood clusters,
  #there were 0 cases of a specific offense.
violent_crime_by_neigh_cluster2$`ASSAULT W/DANGEROUS WEAPON`[
  is.na(violent_crime_by_neigh_cluster2$`ASSAULT W/DANGEROUS WEAPON`)] <- 0
violent_crime_by_neigh_cluster2$ROBBERY[
  is.na(violent_crime_by_neigh_cluster2$ROBBERY)] <- 0
violent_crime_by_neigh_cluster2$`SEX ABUSE`[
  is.na(violent_crime_by_neigh_cluster2$`SEX ABUSE`)] <- 0
violent_crime_by_neigh_cluster2$HOMICIDE[
  is.na(violent_crime_by_neigh_cluster2$HOMICIDE)] <- 0

#Want a column name with just the number of a
  #cluster (e.g., 1 instead of Cluster 1)
violent_crime_by_neigh_cluster2[
  c('Cluster_word', 'Cluster_num')] <- 
  str_split_fixed(violent_crime_by_neigh_cluster2$NEIGHBORHOOD_CLUSTER, ' ', 2)
violent_crime_by_neigh_cluster2$Cluster_word <- NULL
```

```{r}
#| warning: false
#| message: false
#Creating new number of robberies factor variable
  #that contains ranges
violent_crime_by_neigh_cluster2$`Number of robberies` <- NA
violent_crime_by_neigh_cluster2$`Number of robberies`[
  violent_crime_by_neigh_cluster2$ROBBERY >= 0 & 
    violent_crime_by_neigh_cluster2$ROBBERY < 40
  ] <- "0-39"
violent_crime_by_neigh_cluster2$`Number of robberies`[
  violent_crime_by_neigh_cluster2$ROBBERY >= 40 & 
    violent_crime_by_neigh_cluster2$ROBBERY < 80
  ] <- "40-79"
violent_crime_by_neigh_cluster2$`Number of robberies`[
  violent_crime_by_neigh_cluster2$ROBBERY >= 80 & 
    violent_crime_by_neigh_cluster2$ROBBERY < 120
  ] <- "80-119"
violent_crime_by_neigh_cluster2$`Number of robberies`[
  violent_crime_by_neigh_cluster2$ROBBERY >= 120 & 
    violent_crime_by_neigh_cluster2$ROBBERY < 160
  ] <- "120+"

violent_crime_by_neigh_cluster2$`Number of robberies` <- 
  factor(violent_crime_by_neigh_cluster2$`Number of robberies`,
  levels=c("0-39","40-79", "80-119", "120+"))
```

# First graph

```{r}
#| warning: false
#| message: false
#Adding first graph
graph1 <- ggplot(data = violent_crime_by_ward, 
#Aesthetic options: x, y, count
  aes(x=factor(WARD), y=count, 
    fill = factor(WARD))) +
#Non-aesthetic options: stat, color
  geom_bar(stat = "identity", color = "black") +
#Aesthetic option: label
#Non-aesthetic options: vjust, color, alpha
  geom_text(aes(label = count), vjust = 1.5, 
    colour = "black", alpha = 0.8) +
  xlab("Ward") +
  scale_x_discrete(
    limits = factor(c(7,8,5,1,6,2,4,3))) +
  ylab("Number of violent crimes in 2021") +
  scale_y_continuous(breaks=seq(0,800,by=200)) +
  labs(
    title = "Number of violent crimes in each of the 8 wards",
    subtitle = "Violent crime was most common in Wards 7 and 8",
    caption = "Source: Open Data DC (2021)",
    fill = "Ward") +
  theme_classic()
graph1
```

This bar graph shows the number of violent crimes that happened in each of the eight wards of Washington, DC in 2021. The ward where the most violent crimes happened was Ward 7 (917 cases) and the ward where the fewest violent crimes happened was Ward 3 (79). In total, there were 4,106 violent crimes in 2021.

Wards 7 and 8 are high-poverty and have large populations of African Americans. The bar graph helps explain why residents of those wards are especially worried about violent crime.

Note that Open Data DC categorizes homicides, sex abuses, assaults with dangerous weapons, and robberies as violent crimes. Meanwhile, Open Data DC categorizes arson, burglaries, theft with auto, and theft/other as property crimes. This may be a different way of defining violent crime than is used by other researchers but it is the conceptualization that I am using. (There were 24,208 property crimes in 2021, which is not shown in the graph.)
